Most people are familiar with grammar in the context of natural language.
In English class, for example, we might learn that we shouldn't end a sentence with a preposition.
This is a \textit{grammatical rule} of English.
In broad strokes, grammar refers to the rules that govern what constitutes a well-formed sentence in a languange.
It's the concern of grammar to determine that \enquote{I taught the car in the refrigerator} is well-formed, whereas \enquote{I car teach refrigerator in there} is not.
It's \textit{not} the concern of grammar to determine that the first sentence, despite being grammatically correct, has no reasonable interpretation in English.

The grammar of a programming language determines what constitutes a well-formed \textit{program} in that language.
Due to the precision of programming languages, these tend to be called \textbf{formal grammars}.
In OCaml, the program:
%
\begin{ocaml}
  let f x = x + 1
\end{ocaml}
%
is well-formed, but the program:
%
\begin{ocaml}
  f let x = x 1 +
\end{ocaml}
%
is not.
As with natural language, the grammar of a programming language is not concerned with the \textit{meaning} of programs, just their well-formedness, e.g., the program:
%
\begin{ocaml}
  let omega x = x x
\end{ocaml}
%
is well-formed, but isn't well-typed.

A good programming language has a clear specification of its grammar.
The grammar of OCaml, for example, is given in its entirety in \href{https://ocaml.org/manual/5.3/expr.html}{The OCaml Manual}.
After going though this chapter, you should be able to read the specification given there.\footnote{
In order to not be entirely biased, I'll note that Python also has a complete specification of its grammar in \href{https://docs.python.org/3/reference/grammar.html}{The Python Manual}.}

\section{A Thought Experiment}

How do we know that a sentence is grammatically correct?
Let's try to break down the cognitive process of this determination.
Consider the following English sentence.

\begin{align*}
  \bnft{the} \ \bnft{cow} \ \bnft{jumped} \ \bnft{over} \ \bnft{the} \ \bnft{moon}
\end{align*}
%
We might first recognize that each word is a particular part of speech (noun, verb, preposition, etc.) and that, with regards to grammar, \textit{only the part of speech counts}; that is, it doesn't matter \textit{what} noun we use, just that we use a noun in the correct position in our sentence.
We represent this step of the cognitive process by replacing each word with an abstract symbol that stands in for its part of speech.

\begin{align*}
  \bnfnt{article} \ \bnfnt{noun} \ \bnfnt{verb} \ \bnfnt{prep} \ \bnfnt{article} \ \bnfnt{noun}
\end{align*}
%
After this, there are some familiar patterns: $\bnfnt{article} \ \bnfnt{noun}$ is the determination or quantification of an object, e.g., \enquote{the house} or \enquote{a car.}
In other words, this pattern represents a \textit{noun phrase} or \textit{nominal phrase}.
We mentally group these forms, and this group can be represented by a new symbol.

\begin{align*}
  \bnfnt{noun-phrase} \ \bnfnt{verb} \ \bnfnt{prep} \ \bnfnt{noun-phrase}
\end{align*}
%
Another pattern we might recognize: a proposition followed by a noun phrase is a single unit, e.g., \enquote{over the moon}, \enquote{through the woods}, or \enquote{behind the wall.}
These are called \textit{prepositional phrases} and can, again, be represented by a new symbol.

\begin{align*}
  \bnfnt{noun-phrase} \ \bnfnt{verb} \ \bnfnt{prep-phrase}
\end{align*}
%
Prepositional phrases modify verbs, creating a \textit{verb phrase} (e.g., \enquote{ran to the car}) leaving us with something like:

\begin{align*}
  \bnfnt{noun-phrase} \ \bnfnt{verb-phrase}
\end{align*}
%
which we should finally recognized as the canonical structure of a well-formed sentence: \textit{thing does}.

The point of this though of experiment is to emphasize the following: there are general structures that are allowed by a grammar, and these structures can be \textit{instantiated} at particular words in the language.
The rules that define these structures are often heirarchical; a sentence is made of parts (e.g, a noun phrase and a verb phrase) and those constituent parts are made of smaller parts (e.g., an article and a noun) and so on until we reach units of our language, the particular words.
Demonstrating that a sentence is grammatically correct means demonstrating that it fits into this heirarchy of allowed structures.

Next, an observation: if we reverse the order of steps we took above, and break up some of the steps we did in parallel, we get a \textit{proof} that our sentence is grammatically correct (\Cref{fig:cow-deriv}).
We call this a \textbf{derivation}.
Each line corresponds to the application of a grammatical rule.
For example, the fourth line follows from the third by way of applying the grammatical rule that a prepositional phrase can be made up of a preposition followed by a verb phrase.
But more importantly, if we squint, we can start to see the tree-like hierarchical structure embodied in this derivation, a structure we call a \textbf{parse tree} (\Cref{fig:cow-tree}).

\begin{figure}
  \begin{align*}
    & \bnfnt{sentence} \\
    & \bnfnt{noun-phrase} \ \bnfnt{verb-phrase} \\
    & \bnfnt{noun-phrase} \ \bnfnt{verb} \ \bnfnt{prep-phrase} \\
    & \bnfnt{noun-phrase} \ \bnfnt{verb} \ \bnfnt{prep} \ \bnfnt{noun-phrase} \\
    & \bnfnt{noun-phrase} \ \bnfnt{verb} \ \bnfnt{prep} \ \bnfnt{article} \ \bnfnt{noun} \\
    & \bnfnt{article} \ \bnfnt{noun} \ \bnfnt{verb} \ \bnfnt{prep} \ \bnfnt{article} \ \bnfnt{noun} \\
    & \bnft{the} \ \bnfnt{noun} \ \bnfnt{verb} \ \bnfnt{prep} \ \bnfnt{article} \ \bnfnt{noun} \\
    & \bnft{the} \ \bnft{cow} \ \bnfnt{verb} \ \bnfnt{prep} \ \bnfnt{article} \ \bnfnt{noun} \\
    & \bnft{the} \ \bnft{cow} \ \bnft{jumped} \ \bnfnt{prep} \ \bnfnt{article} \ \bnfnt{noun} \\
    & \bnft{the} \ \bnft{cow} \ \bnft{jumped} \ \bnft{over} \ \bnfnt{article} \ \bnfnt{noun} \\
    & \bnft{the} \ \bnft{cow} \ \bnft{jumped} \ \bnft{over} \ \bnft{the} \ \bnfnt{noun} \\
    & \bnft{the} \ \bnft{cow} \ \bnft{jumped} \ \bnft{over} \ \bnft{the} \ \bnft{moon}
  \end{align*}
  \caption{Derivation of $\bnft{the} \ \bnft{cow} \ \bnft{jumped} \ \bnft{over} \ \bnft{the} \ \bnft{moon}$ in the grammar in \Cref{fig:simple-grammar}}
  \label{fig:cow-deriv}
\end{figure}

\begin{figure}
  \begin{forest}
    [$\bnfnt{sentence}$
      [$\bnfnt{noun-phrase}$
        [$\bnfnt{article}$
          [$\bnft{the}$]
        ]
        [$\bnfnt{noun}$
          [$\bnft{cow}$]
        ]
      ]
      [$\bnfnt{verb-phrase}$
        [$\bnfnt{verb}$
          [$\bnft{jumped}$]
        ]
        [$\bnfnt{prep-phrase}$
          [$\bnfnt{prep}$
            [$\bnft{over}$]
          ]
          [$\bnfnt{noun-phrase}$
            [$\bnfnt{article}$
              [$\bnft{the}$]
            ]
            [$\bnfnt{noun}$
              [$\bnft{moon}$]
            ]
          ]
        ]
      ]
    ]
  \end{forest}
  \centering
  \caption{Parse tree for $\bnft{the} \ \bnft{cow} \ \bnft{jumped} \ \bnft{over} \ \bnft{the} \ \bnft{moon}$ in the grammar in \Cref{fig:simple-grammar}}
  \label{fig:cow-tree}
\end{figure}

\section{Backus-Naur Form}

A formal grammar is defined over a fixed collection of symbols.
This collection is divided into two disjoint groups: the \textbf{terminal symbols} and the \textbf{non-terminal} symbols.
We denote nonterminal symbols using notation of the form $\bnfnt{non-terminal}$.
We denote terminal symbols in red typewriter font, e.g., $\bnft{terminal}$.

\begin{remark}
  We almost never state outright what our underlying symbols are.
  We can determine the terminal and non-terminal symbols of a grammar by looking at its specification, assuming that every symbol must appear at least once in the specification.
\end{remark}
%
In the derivation in \Cref{fig:cow-deriv}, we built a sequence of not-quite sentences, until the very last one, which was the sentence we wanted to prove grammatically correct.
These possibly-not-quite sentences are called \textbf{sentential forms}.

\begin{definition}
  A \textbf{sentential form} is a sequence of symbols (terminal or non-terminal).
  A \textbf{sentence} is a sequence of terminal symbols.
  We will use the symbol $\bnfempty$ to refer to the empty sentential form.
\end{definition}

\begin{remark}
  We denote sentential forms using whitespace as a delimiter.
  But it's important to recognize that \textit{this is just notation.}
  We're not interested in low-level concerns like whitespace when studying formal grammar.
  It may be useful to think of a sentential form as a \textit{list} of symbols.
\end{remark}

\begin{example}
  The following are sentential forms that appear above:
  \begin{align*}
    & \bnfnt{sentence} \\
    & \bnfnt{noun-phrase} \ \bnfnt{verb} \ \bnfnt{prep} \ \bnfnt{article} \ \bnfnt{noun} \\
    & \bnft{the} \ \bnft{cow} \ \bnft{jumped} \ \bnft{over} \ \bnft{the} \ \bnft{moon}
  \end{align*}
  but only the last one is a sentence.
\end{example}
%
In the process of constructing sentential forms, we replaced a non-terminal symbol with a sentential form, e.g., we replaced $\bnfnt{noun-phrase}$ with $\bnfnt{article} \ \bnfnt{noun}$.
These replacements are specified using \textbf{production rules}.

\begin{definition}
  A \textbf{production rule} is a nonterminal symbol together with a sentential form.
  Production rules are denoted using equations of the form
  \begin{align*}
    \bnfnt{non-terminal} \bnfeq \textsf{SENTENTIAL-FORM}
  \end{align*}
  where the left-hand side is the nonterminal symbol and the right-hand side is the sentential form.
\end{definition}
%
We read a production rule as: \enquote{the non-terminal symbol on the left-hand side can be replaced with the sentential form on the right hand side.}
%
\begin{example}
  We can express that a prepositional phrase can be made up of a preposition followed by a noun-phrase using the following production rule:
  \begin{align*}
    \bnfnt{prep-phrase} \bnfeq \bnfnt{prep} \ \bnfnt{noun-phrase}
  \end{align*}
\end{example}
%
\textbf{Backus-Naur Form (BNF) specifications} are used to describe \textbf{context-free grammars}, which form a class of formal grammars that are expressive enough to capture the syntax of programming languages.
A BNF grammar is determined by what replacements we're allowed to perform and what kind of things we're trying to derive.

\begin{definition}
A \textbf{Backus-Naur Form (BNF) specification} is a collection of production rules, together with a designated symbol called the \textbf{starting symbol}.
If the start symbols is not explicitely given, it is taken to be the left-hand side of the \textit{first} rule appearing in the specification.
\end{definition}

\begin{figure}
\begin{align*}
\bnfnt{sentence} &\bnfeq \bnfnt{noun-phrase} \ \bnfnt{verb-phrase} \\
\bnfnt{verb-phrase} &\bnfeq \bnfnt{verb} \ \bnfnt{prep-phrase} \\
\bnfnt{verb-phrase} &\bnfeq \bnfnt{verb} \\
\bnfnt{prep-phrase} &\bnfeq \bnfnt{prep} \ \bnfnt{noun-phrase} \\
\bnfnt{noun-phrase} &\bnfeq \bnfnt{article} \ \bnfnt{noun} \\
\bnfnt{article} &\bnfeq \bnft{the} \\
\bnfnt{noun} &\bnfeq \bnft{cow} \\
\bnfnt{noun} &\bnfeq \bnft{moon} \\
\bnfnt{verb} &\bnfeq \bnft{jumped} \\
\bnfnt{prep} &\bnfeq \bnft{over}
\end{align*}
\caption{BNF specification for a simple grammar based on English}
\label{fig:simple-grammar}
\end{figure}

\begin{example}
  \Cref{fig:simple-grammar} has a BNF specification for the underlying grammar of the previous section.
  The nonterminal symbols of this grammar are: $\bnfnt{sentence}$, $\bnfnt{noun-phrase}$, $\bnfnt{verb-phrase}$, $\bnfnt{verb}$, $\bnfnt{prep}$, $\bnfnt{prep-phrase}$, $\bnfnt{article}$, and $\bnfnt{noun}$.
  The terminal symbols of this grammar are: $\bnft{the}$, $\bnft{cow}$, $\bnft{moon}, $\bnft{jumped}, and $\bnft{over}$.
  The nonterminal symbol $\bnfnt{sentence}$ is our starting symbol (because it appears as the left-hand side of the first rule).
  This is, of course, not a complete or accurate model of English grammar (such a model would be incredibly complicated).
\end{example}

\begin{remark}
  John Backus and Peter Naur are computer scientists who formalized this meta-syntax in the process of developing and implementing the programming language \href{https://dl.acm.org/doi/pdf/10.1145/366193.366201}{ALGOL 60}.
\end{remark}
%
As \Cref{fig:simple-grammar} indicates, it's possible for a nonterminal symbol to have multiple associated production rules.
This is very common so it has it's own syntax.
%
\begin{definition}
We write
%
\begin{align*}
\bnfnt{non-terminal} \bnfeq \textsf{SENTENTIAL-FORM}_1 \bnfalt \dots \bnfalt \textsf{SENTENTIAL-FORM}_k
\end{align*}
%
as shorthand for
%
\begin{align*}
\bnfnt{non-terminal} &\bnfeq \textsf{SENTENTIAL-FORM}_1 \\
&\vdots \\
\bnfnt{non-terminal} &\bnfeq \textsf{SENTENTIAL-FORM}_k
\end{align*}
%
The multiple forms of a given nonterminal symbol are often called \textbf{alternatives}, and this shorthand is often called \textbf{alternative syntax}.
\end{definition}
%
It's not difficult to rewrite the grammar in \Cref{fig:simple-grammar} using alternative syntax (\Cref{fig:simple-grammar-alt}).
The last piece of our thought experiment that we need to formalize is the \textit{proof} of grammatical correctness.

\begin{figure}
\begin{align*}
\bnfnt{sentence} &\bnfeq \bnfnt{noun-phrase} \ \bnfnt{verb-phrase} \\
\bnfnt{verb-phrase} &\bnfeq \bnfnt{verb} \ \bnfnt{prep-phrase} \bnfalt \bnfnt{verb} \\
\bnfnt{prep-phrase} &\bnfeq \bnfnt{prep} \ \bnfnt{noun-phrase} \\
\bnfnt{noun-phrase} &\bnfeq \bnfnt{article} \ \bnfnt{noun} \\
\bnfnt{article} &\bnfeq \bnft{the} \\
\bnfnt{noun} &\bnfeq \bnft{cow} \bnfalt \bnft{moon} \\
\bnfnt{verb} &\bnfeq \bnft{jumped} \\
\bnfnt{prep} &\bnfeq \bnft{over}
\end{align*}
\caption{The grammar in \Cref{fig:simple-grammar} rewritten with alternative syntax}
\label{fig:simple-grammar-alt}
\end{figure}

\begin{definition}
  \label{def:deriv}
  A \textbf{derivation} of a sentence $S$ in a BNF grammar $\mathcal G$ is a sequence of sentential forms $S_1, \dots, S_k$ with the following properties:
  \begin{itemize}
  \item $S_1$ is the starting symbol of $\mathcal G$;
  \item $S_k = S$;
  \item for every index $i$ satisfying $i > 1$, the sentential form $S_i$ is the result of replacing a single nonterminal symbol in $S_{i - 1}$ with a sentential form, according to a production rule of $\mathcal G$.
  \end{itemize}
  We say that a grammar \textbf{recognizes} a sentence $S$ if there is a derivation of $S$ in the grammar.
\end{definition}
%
\begin{example}
  \label{ex:moon-deriv}
  The derivation in \Cref{fig:cow-deriv} is a derivation is the sense of \Cref{def:deriv}.
  We can also derive the sentence $\bnft{the} \ \bnft{moon} \ \bnft{jumped}$ as follows:
  %
  \begin{align*}
    & \bnfnt{sentence} \\
    & \bnfnt{noun-phrase} \ \bnfnt{verb-phrase} \\
    & \bnfnt{article} \ \bnfnt{noun} \ \bnfnt{verb-phrase} \\
    & \bnft{the} \ \bnfnt{noun} \ \bnfnt{verb-phrase} \\
    & \bnft{the} \ \bnft{moon} \ \bnfnt{verb-phrase} \\
    & \bnft{the} \ \bnft{moon} \ \bnfnt{verb} \\
    & \bnft{the} \ \bnft{moon} \ \bnft{jumped}
  \end{align*}
\end{example}

\begin{exercise}
  For the derivation in \Cref{ex:moon-deriv}, determine which production rule is applied in each line of the derivation.
\end{exercise}
%
A sentence is not guaranteed to have a \textit{unique} derivation, but there a form of derivation we'll single out for reasons which will become more clear at the end of this section.

\begin{definition}
  A \textbf{leftmost derivation} is one in which the leftmost nonterminal symbol is expanded in each step.
\end{definition}
%
The derivation given \Cref{ex:moon-deriv} is a leftmost derivation, whereas the derivation in \Cref{fig:cow-deriv} is not.

\begin{exercise}
  Write a leftmost derivation of the sentence
  \begin{align*}
    \bnft{the} \ \bnft{cow} \ \bnft{jumped} \ \bnft{over} \ \bnft{the} \ \bnft{moon}
  \end{align*}
  in the grammar in \Cref{fig:simple-grammar}.
\end{exercise}
%
The theme of this section has been that grammars imbue sentences with hierarchical structure.
We represent this hierarchical structure using \textbf{parse trees}.

\begin{definition}
  \label{def:parse-tree}
  A \textbf{parse tree} for a sentence $S$ recognized by a grammar $\mathcal G$ is a tree $T$ with the following properties:

  \begin{itemize}
  \item The values of $T$ are symbols of $\mathcal G$;
  \item $\troot T$ is the starting symbol of $\mathcal G$;
  \item For every node of $T$ of the form $\tnode(s, T_1, \dots, T_k)$
    \begin{align*}
      s \bnfeq \troot {T_1} \ \dots \ \troot{T_k}
    \end{align*}
    is a production rule of $\mathcal G$;
  \item The terminal symbols at the leaves of $T$, in order from left to right form the sentence $S$.
  \end{itemize}
  %
  Note that the value at a leaf $l$ may be a nonterminal symbol in the case that
  %
  \begin{align*}
    \troot l \bnfeq \bnfempty
  \end{align*}
  %
  is a production rule of $\mathcal G$.
\end{definition}

\begin{example}
  The tree in \Cref{fig:cow-tree} is a parse tree in the sense of \label{def:parse-tree}.
  The following is a parse tree for the sentence $\bnft{the} \ \bnft{moon} \ \bnft{jumped}$ in the same grammar:
  \begin{center}
    \begin{forest}
      [$\bnfnt{sentence}$
        [$\bnfnt{noun-phrase}$
          [$\bnfnt{article}$
            [$\bnft{the}$]
          ]
          [$\bnfnt{noun}$
            [$\bnft{moon}$]
          ]
        ]
        [$\bnfnt{verb-phrase}$
          [$\bnfnt{verb}$
            [$\bnft{jumped}$]
          ]
        ]
      ]
    \end{forest}
\end{center}
\end{example}
%
If a sentence $S$ has a derivation in a grammar $\mathcal G$, then it also has a parse tree, and vice versa.
Multiple derivations may correspond to the same parse tree, but to every parse tree there is a unique corresponding leftmost derivation; a leftmost derivation corresponds to a pre-order traversal of a parse tree.

\begin{theorem}
  \label{thm:tree-leftmost}
  There is a one-to-one correspondence between parse trees and leftmost derivations.
  In particular, a sentence $S$ has exactly one parse tree in $\mathcal G$ if and only if it has a exactly one leftmost derivation in $\mathcal G$.
\end{theorem}

\begin{exercise}
  \textbf{(Challenge)} Prove \Cref{thm:tree-leftmost}.
\end{exercise}

\begin{remark}
  In addition to context-free grammars, there are also context-\textit{sensitive} grammars.
  For context-sensitive grammars, production rules can transform \textit{sequences of} symbols, i.e., the left-hand side of a production rule is not required to be a single nonterminal symbol.
\end{remark}

\section{Examples}

Despite all the abstractions and formalisms, formal grammars are suppose to be fairly straigtfoward.
The easiest way to grok them is by example.
In this section, we'll look at a couple more complicated examples based on programming languages.

\subsection*{An Imperative Language}

\begin{figure}
  \begin{align*}
    \bnfnt{program} &\bnfeq \bnfnt{stmts} \\
    \bnfnt{stmts} &\bnfeq \bnfempty \bnfalt \bnfnt{stmt} \ \bnft{;} \ \bnfnt{stmts} \\
    \bnfnt{stmt} &\bnfeq \bnfnt{var} \ \bnft{=} \ \bnfnt{expr} \\
    \bnfnt{expr} &\bnfeq
    \bnfnt{term} \bnfalt
    \bnfnt{term} \ \bnft{+} \ \bnfnt{term} \bnfalt
    \bnfnt{term} \ \bnft{-} \ \bnfnt{term} \\
    \bnfnt{term} &\bnfeq \bnfnt{var} \bnfalt \bnfnt{num} \\
    \bnfnt{num} &\bnfeq \bnft{1} \bnfalt \bnft{2} \bnfalt \bnft{3} \\
    \bnfnt{var} &\bnfeq \bnft{a} \bnfalt \bnft{b} \bnfalt \bnft{c} \bnfalt \bnft{d}
  \end{align*}
  \caption{BNF specification for a toy imperative programming language}
  \label{fig:line-pl}
\end{figure}

\Cref{fig:line-pl} has a BNF specification for a toy imperative programming language which, in essense, consists of what are called \textit{straight-line programs}.
In English, we would read this specification as:
\begin{itemize}
\item a \textit{program} is made up of \textit{statements};
\item a collection of \textit{statements} is empty or is a single \textit{statement}, followed a semicolon, followed by a collection of \textit{statements};
\item a statement is a \textit{variable}, followed by an equals sign, followed by a \textit{expression};
\item and so on\dots
\end{itemize}
%
The second rule highlights an interesting feature of BNF specifications: production rules are allowed to be \textit{recursive}.
The production rule for $\bnfnt{stmts}$ allows us to expand $\bnfnt{stmts}$ to a sentential form that \textit{contains} the non-terminal symbol $\bnfnt{stmts}$.
This means the above grammar recognizes infinitely many sentences.

Consider the following program in this language.

\begin{lstlisting}
a = 1;
a = a + 2;
b = a;
\end{lstlisting}
%
To verify that this program is recognized by the above grammar, we can write a (leftmost) derivation of it (\Cref{fig:line-deriv}).

\begin{figure}
\begin{align*}
& \bnfnt{program} \\
& \bnfnt{stmts} \\
& \bnfnt{stmt} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnfnt{var} \ \bnft{=} \ \bnfnt{expr} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnfnt{expr} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnfnt{term} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnfnt{num} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnfnt{stmt} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnfnt{var} \ \bnft{=} \ \bnfnt{expr} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnfnt{expr} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnfnt{term} \ \bnft{+} \ \bnfnt{term} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnfnt{var} \ \bnft{+} \ \bnfnt{term} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnft{a} \ \bnft{+} \ \bnfnt{term} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnft{a} \ \bnft{+} \ \bnfnt{num} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnft{a} \ \bnft{+} \ \bnft{2} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnft{a} \ \bnft{+} \ \bnft{2} \ \bnft{;} \ \bnfnt{stmt} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnft{a} \ \bnft{+} \ \bnft{2} \ \bnft{;} \ \bnfnt{var} \ \bnft{=} \ \bnfnt{expr} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnft{a} \ \bnft{+} \ \bnft{2} \ \bnft{;} \ \bnft{b} \ \bnft{=} \ \bnfnt{expr} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnft{a} \ \bnft{+} \ \bnft{2} \ \bnft{;} \ \bnft{b} \ \bnft{=} \ \bnfnt{term} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnft{a} \ \bnft{+} \ \bnft{2} \ \bnft{;} \ \bnft{b} \ \bnft{=} \ \bnfnt{var} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnft{a} \ \bnft{+} \ \bnft{2} \ \bnft{;} \ \bnft{b} \ \bnft{=} \ \bnft{a} \ \bnft{;} \ \bnfnt{stmts} \\
& \bnft{a} \ \bnft{=} \ \bnft{1} \ \bnft{;} \ \bnft{a} \ \bnft{=} \ \bnft{a} \ \bnft{+} \ \bnft{2} \ \bnft{;} \ \bnft{b} \ \bnft{=} \ \bnft{a} \ \bnft{;}
\end{align*}
\caption{Example derivation in the grammar in \Cref{fig:line-pl}}
\label{fig:line-deriv}
\end{figure}
%

\begin{remark}
As a reminder, we're not interested in low-level syntactic concerns like whitespace when we consider whether or not a sentence is recognized by a grammar.
The choice to present the program above in three lines was for readability, and the choice to present it in a single line in the derivation was for convenience.
\end{remark}
%
Alternatively, we could construct a parse for the given program (\Cref{fig:line-tree}).

\begin{figure}
  \resizebox{\textwidth}{!}{%
    \begin{forest}
      [$\bnfnt{program}$
        [$\bnfnt{stmts}$
          [$\bnfnt{stmt}$
            [$\bnfnt{var}$
              [$\bnft{a}$]
            ]
            [$\bnft{=}$]
            [$\bnfnt{expr}$
              [$\bnfnt{term}$
                [$\bnfnt{num}$
                  [$\bnft{1}$]
                ]
              ]
            ]
          ]
          [$\bnft{;}$]
          [$\bnfnt{stmts}$
            [$\bnfnt{stmt}$
              [$\bnfnt{var}$
                [$\bnft{a}$]
              ]
              [$\bnft{=}$]
              [$\bnfnt{expr}$
                [$\bnfnt{term}$
                  [$\bnfnt{var}$
                    [$\bnft{a}$]
                  ]
                ]
                [$\bnft{+}$]
                [$\bnfnt{term}$
                  [$\bnfnt{num}$
                    [$\bnft{2}$]
                  ]
                ]
              ]
            ]
            [$\bnft{;}$]
            [$\bnfnt{stmts}$
              [$\bnfnt{stmt}$
                [$\bnfnt{var}$
                  [$\bnft{b}$]
                ]
                [$\bnft{=}$]
                [$\bnfnt{expr}$
                  [$\bnfnt{term}$
                    [$\bnfnt{var}$
                      [$\bnft{a}$]
                    ]
                  ]
                ]
              ]
              [$\bnft{;}$]
              [$\bnfnt{stmts}$]
            ]
          ]
        ]
      ]
    \end{forest}
  }
  \centering
  \caption{
    Example parse tree for the grammar in \Cref{fig:line-pl}.
  }
  \label{fig:line-tree}
\end{figure}

\begin{exercise}
Verify that
\begin{lstlisting}
a = a + a;
b = b;
\end{lstlisting}
is recognized by the above grammar by giving both a derivation and a parse tree.
\end{exercise}

\subsection*{Arithmetic Expressions}

\Cref{fig:arith-simple} has a first attempt at a BNF specification for arithmetic expressions.

\begin{figure}
  \begin{align*}
    \bnfnt{expr} &\bnfeq
    \bnfnt{num} \bnfalt
    \bnfnt{expr} \ \bnfnt{op} \ \bnfnt{expr} \bnfalt
    \bnft{(} \ \bnfnt{expr} \ \bnft{)} \\
    \bnfnt{op} &\bnfeq \bnft{+} \bnfalt \bnft{-} \bnfalt \bnft{*} \bnfalt \bnft{/} \\
    \bnfnt{num} &\bnfeq \bnft{1} \bnfalt \bnft{2} \bnfalt \bnft{3}
  \end{align*}
  \caption{First attempt at a BNF specification for arithmetic expressions}
  \label{fig:arith-simple}
\end{figure}
%
\begin{exercise}
Give a derivation and parse tree for the following sentence:
%
\begin{align*}
\bnft{(} \ \bnft{1} \ \bnft{+} \ \bnft{2} \ \bnft{)} \ \bnft{*} \ \bnft{3}
\end{align*}
using the grammar in \Cref{fig:arith-simple}.
\end{exercise}
%
This grammar seems, ignoring obvious issues like having only three possible numbers, to be a reasonable enough specification: an expression is either a number or a pair of expressions with an operator between them.
Note that the recursive nature of the first production rule means that this grammar recognizes infinitely many sentences.
But this grammar has a curious feature: \textit{there are sentences which have multiple parse trees.}
Consider the sentence \codel{1 - 2 - 3}.
One the one hand, we might thing that this expression says \enquote{subtract 2 from 1, and then subtract 3.}
This reading is witnessed by the following parse tree:

\begin{center}
  \begin{forest}
    [$\bnfnt{expr}$
      [$\bnfnt{expr}$
        [$\bnfnt{num}$
          [$\bnft{1}$]
        ]
        [$\bnft{-}$]
        [$\bnfnt{num}$
          [$\bnft{2}$]
        ]
      ]
      [$\bnft{-}$]
      [$\bnfnt{num}$
        [$\bnft{3}$]
      ]
    ]
  \end{forest}
\end{center}
%
But there's nothing in the structure of the expression itself which keeps us from intepreting it as \enquote{subtract from 1 the result of subtracting 3 from 2} which is witnessed by this alternative parse tree:

\begin{center}
  \begin{forest}
    [$\bnfnt{expr}$
      [$\bnfnt{num}$
        [$\bnft{1}$]
      ]
      [$\bnft{-}$]
      [$\bnfnt{expr}$
        [$\bnfnt{num}$
          [$\bnft{2}$]
        ]
        [$\bnft{-}$]
        [$\bnfnt{num}$
          [$\bnft{3}$]
        ]
      ]
    ]
  \end{forest}
\end{center}
%
This ambiguity matters for the design of programming languages because it can change the \textit{meaning} of the program itself; the first reading gives us the value $-4$ whereas the latter gives us $2$.
If we want to allow the programmer to write down an expression like \codel{1 - 2 - 3}, then we have to be clear about what the intended ouput should be.
More generally, we'd like to ensure that our grammar is \textit{unambiguous} so there's no question as to how to read what the programmer has written.
This will be our focus for the next section.

\section{Grammatical Ambiguity}

As participants of language, we are no strangers to grammatical ambiguity.
Take, for instance, the following sentence:\footnote{This example is taken from the Wikipedia page on \href{https://en.wikipedia.org/wiki/Syntactic_ambiguity}{Syntactic ambiguity}.}

\begin{align*}
  \bnft{John saw the man on the mountain with the telescope}
\end{align*}
%
Was John using the telescope to see the man on the mountain?
Was the man carrying the telescope?
Are there multiple mountains, one of which has a telescope on it?
The ambiguity comes from it not being clear \textit{which} hierarchical structure should scaffold the sentence.
\Cref{fig:ambiguous-tree-1} has one possible parse tree for this sentence in grammar based on English (\Cref{fig:john-grammar}).
The prepositional phrase \enquote{with the telescope} is grouped with the verb phrase starting with \enquote{saw}, indicating that John was \textit{using} the telescope.
\Cref{fig:ambiguous-tree-2} has an alternative parse tree for the same sentence.
For this parse tree, the prepositional phrase \enquote{with the telescope} is grouped with the noun phrase \enquote{the mountain} indicating that there is a telescope on the mountain itself.

\begin{figure}
  \begin{forest}
    [$\bnfnt{s}$
      [$\bnfnt{np}$
        [$\bnfnt{n}$
          [$\bnft{John}$]
        ]
      ]
      [$\bnfnt{vp}$
        [$\bnfnt{vp}$
          [$\bnfnt{v}$
            [$\bnft{saw}$]
          ]
          [$\bnfnt{np}$
            [$\bnfnt{np}$
              [$\bnfnt{a}$
                [$\bnft{the}$]
              ]
              [$\bnfnt{n}$
                [$\bnft{man}$]
              ]
            ]
            [$\bnfnt{pp}$
              [$\bnfnt{p}$
                [$\bnft{on}$]
              ]
              [$\bnfnt{np}$
                [$\bnfnt{a}$
                  [$\bnft{the}$]
                ]
                [$\bnfnt{n}$
                  [$\bnft{mountain}$]
                ]
              ]
            ]
          ]
        ]
        [$\bnfnt{pp}$
          [$\bnfnt{p}$
            [$\bnft{with}$]
          ]
          [$\bnfnt{np}$
            [$\bnfnt{a}$
              [$\bnft{the}$]
            ]
            [$\bnfnt{n}$
              [$\bnft{telescope}$]
            ]
          ]
        ]
      ]
    ]
  \end{forest}
  \centering
  \caption{Parse tree of an ambiguous sentence for the grammar in \Cref{fig:john-grammar}}
  \label{fig:ambiguous-tree-1}
\end{figure}

\begin{figure}
  \begin{forest}
    [$\bnfnt{s}$
      [$\bnfnt{np}$
        [$\bnfnt{n}$
          [$\bnft{John}$]
        ]
      ]
      [$\bnfnt{vp}$
        [$\bnfnt{v}$
          [$\bnft{saw}$]
        ]
        [$\bnfnt{np}$
          [$\bnfnt{np}$
            [$\bnfnt{a}$
              [$\bnft{the}$]
            ]
            [$\bnfnt{n}$
              [$\bnft{man}$]
            ]
          ]
          [$\bnfnt{pp}$
            [$\bnfnt{p}$
              [$\bnft{on}$]
            ]
            [$\bnfnt{np}$
              [$\bnfnt{np}$
                [$\bnfnt{a}$
                  [$\bnft{the}$]
                ]
                [$\bnfnt{n}$
                  [$\bnft{mountain}$]
                ]
              ]
              [$\bnfnt{pp}$
                [$\bnfnt{p}$
                  [$\bnft{with}$]
                ]
                [$\bnfnt{np}$
                  [$\bnfnt{a}$
                    [$\bnft{the}$]
                  ]
                  [$\bnfnt{n}$
                    [$\bnft{telescope}$]
                  ]
                ]
              ]
            ]
          ]
        ]
      ]
    ]
  \end{forest}
  \centering
  \caption{Another parse tree of an ambiguous sentence for the grammar in \Cref{fig:john-grammar}.}
  \label{fig:ambiguous-tree-2}
\end{figure}

\begin{remark}
  We experience language in a linear fashion, either by reading it or hearing it.
  Ifn our interlocutor could \textit{display} the parse tree of their statement (floating eerily in space before our eyes) there would be nothing to say of (grammatical) ambiguity.

  In another light, grammatical ambiguity is related to parentheses.
  If it were standard practice to use parentheses in natural language, we could avoid some grammatical ambiguity.
  For example, the parse tree in \Cref{fig:ambiguous-tree-1} corresponds to:
  %
  \begin{align*}
    \bnft{John saw ( the man on the mountain ) with the telescope}
  \end{align*}
  %
  whereas the parse tree in \Cref{fig:ambiguous-tree-2} corresponds to:
  %
  \begin{align*}
    \bnft{John saw the man on ( the mountain with the telescope )}
  \end{align*}
\end{remark}
%
Ambiguity in natural language is a complex topic, but restricted to formal grammars, ambiguity is a well-defined notion.

\begin{definition}
  A grammar is \textbf{ambiguous} if it recognizes a sentence with (at least) two distinct parse trees.
  Equivalently, it is ambiguous if it recognizes a sentence with (at least) two distinct leftmost derivations.
\end{definition}

\begin{figure}
  \begin{align*}
    \bnfnt{s} &\bnfeq \bnfnt{np} \ \bnfnt{vp} \\
    \bnfnt{vp} &\bnfeq
    \bnfnt{v} \bnfalt
    \bnfnt{v} \ \bnfnt{np} \bnfalt
    \bnfnt{vp} \ \bnfnt{pp} \\
    \bnfnt{np} &\bnfeq
    \bnfnt{n} \bnfalt
    \bnfnt{a} \ \bnfnt{n} \bnfalt
    \bnfnt{np} \ \bnfnt{pp} \\
    \bnfnt{n} &\bnfeq
    \bnft{John} \bnfalt
    \bnft{man} \bnfalt
    \bnft{mountain} \bnfalt
    \bnft{telescope} \\
    \bnfnt{v} &\bnfeq \bnft{saw} \\
    \bnfnt{a} &\bnfeq \bnft{the} \\
    \bnfnt{p} &\bnfeq \bnft{on} \bnfalt \bnft{with}
  \end{align*}
  \caption{BNF grammar that recognizes the ambiguous sentence in this section.}
  \label{fig:john-grammar}
\end{figure}

\begin{example}
  The grammar in \Cref{fig:john-grammar} is ambiguous.
  That is, in this grammar, our ambiguous sentence has multiple parse trees.
  Equivalently, by \Cref{thm:tree-leftmost}, it has multiple leftmost derivations.
  The grammar in \Cref{fig:arith-simple} is also ambiguous by virtue of the examples given in the previous section.
\end{example}

\begin{exercise}
  Give a third distinct parse tree for the \enquote{John} sentence.
\end{exercise}

\begin{exercise}
  Give two (or three) leftmost derivations of the \enquote{John} sentence.
\end{exercise}

\begin{remark}
  In the previous example, along with many of the examples we'll see, it'll be fairly obvious that the grammar is ambiguous.
  As computer scientists, we might think that we could write program to determine if a grammar is ambiguous.
  Unfortunately, this is impossible (not just very difficult, but \textit{impossible}).
  Formally speaking, this is to say that determining if a context-free grammar is ambiguous is \textit{undecidable} \cite{TODO}.
\end{remark}
%
Our next task it to determine how to \textit{avoid} grammatical ambiguity when possible.
But first, \textit{why should we care?}
Natural language is ambiguous and we get along perfectly fine, so why should we go through the trouble of making sure grammars we design for programming languages are unambiguous?
The way I see it, it's a promise that to a programmer: \textit{we never make unspoken assumptions about what you meant when we read your program.}
We try to do this with natural language too, but in communication, if a statement is ambiguous, we can usually just ask our interlocutor what they meant.
We can't do this for a program, so instead we make it \textit{impossible} for a sentence to have multiple readings.\footnote{We see a similar phenomena in legal language, which tends to be grammatically sterile, and no fun to read.}

\subsection{Fixity}

\begin{figure}
  \begin{align*}
    \bnfnt{expr} &\bnfeq
    \bnfnt{num} \bnfalt
    \bnfnt{op} \ \bnfnt{expr} \ \bnfnt{expr} \\
    \bnfnt{op} &\bnfeq \bnft{+} \bnfalt \bnft{-} \bnfalt \bnft{*} \bnfalt \bnft{/} \\
    \bnfnt{num} &\bnfeq \bnft{1} \bnfalt \bnft{2} \bnfalt \bnft{3}
  \end{align*}
  \caption{BNF specification for arithmetic expressions in Polish Notation}
  \label{fig:arith-polish}
\end{figure}

If our only concern is avoiding ambiguity, we can do so by making sure operators appear \textit{before} all their arguments.
\Cref{fig:arith-polish} has a such a grammar for arithmetic expressions.
This is called \textbf{Polish notation}, and always yields an unambiguous grammar, and \textit{without any need for parentheses.}

It's not difficult to then guess what reverse polish notation is: operators appear \textit{after} their arguments.
This is how early calculators like the \href{https://en.wikipedia.org/wiki/Hewlett-Packard_9100A]}{HP 9100A Desktop Calculator} were designed.
If you wanted to calculate something like \codel{(2 + 3) * (4 - 5)}, you would push the operands onto a \textit{stack} and then apply operators to the
top elements of the stack:
%
\begin{align*}
  \emptyset \xlongrightarrow{\codel{2}} [2]
  \xlongrightarrow{\codel{3}} [2, 3]
  \xlongrightarrow{\codel{+}} [5]
  \xlongrightarrow{\codel{4}} [5, 4]
  \xlongrightarrow{\codel{5}} [5, 4, 5]
  \xlongrightarrow{\codel{-}} [5, -1]
  \xlongrightarrow{\codel{*}} [-5]
\end{align*}
%
so that the expression you typed was in reverse Polish notation, in this case:
%
\begin{align*}
  \codel{2 3 + 4 5 - *}.
\end{align*}

\begin{exercise}
  Derive the sentence \codel{+ * 1 * 2 - 3 2 1} using the grammar in \Cref{fig:arith-polish}.
\end{exercise}
%
The benefit of Polish notation is that it is extremely simple, and easy to parse.
The issue with Polish notation is that it's difficult to read.
Imagine having to write code with a prefix form of if-then-else.
Code in OCaml like:
%
\begin{ocaml}
  if n < 10
  then -1
  else if n > 10
  then 1
  else 0
\end{ocaml}
becomes something like:
\begin{lstlisting}
  if_then_else < n 10
    -1
    if_then_else > n 10
      1
      0
\end{lstlisting}
which isn't \textit{so} bad, but is clearly less readable.\footnote{
  There are cases in which the simplicity trumps readability, e.g., for dialects of \href{https://www.forth.com/forth/}{Forth}.
}
This is all to say that ambiguity in expressions is in part due to issues of operator \textbf{fixity}.
%
\begin{definition}
The \textbf{fixity} of an operator refers to where the syntactic components of an operator are placed relative to its arguments.
There are four kinds of operator fixity.
\begin{itemize}
\item
A \textbf{prefix} operator appears before its arguments.
In the expression \codel{-5} the negation operator \codel{-} appears before its argument \codel{5}.
\item
A \textbf{postfix} operator appears after its arguments.
In the expression \codel{10!} the factorial operator \codel{!} appears after its argument \codel{10}.
\item
A \textbf{infix} operator appears between its arguments.
Infix operators are required to be binary.
In the expression \codel{4 + 3} the binary addition operator \codel{+} appears between its arguments \codel{4} and \codel{3}.
\item
A \textbf{mixfix} operator has multiple syntactic components which appear in some combination of before, after, and in between its arguments.
In the expression \codel{if x < 0 then -x else x} the syntactic components \codel{if}, \codel{then}, and \codel{else} appear in various positions with respect to the arguments of the if-then-else operator.
\end{itemize}
\end{definition}
%
The above discussion indicates that if we're willing to accept \textit{only} prefix operators or \textit{only} postfix operators, then it's easy to design unambiguous grammars.
But if we want to contend with multiple kinds of operator fixity, then we'll have to confront the issues of grammatical ambiguity.

\subsection{Parentheses}

\begin{figure}
\begin{align*}
\bnfnt{expr} &\bnfeq
\bnfnt{num} \bnfalt
\bnft{(} \ \bnfnt{expr} \ \bnfnt{op} \ \bnfnt{expr} \bnft{)} \\
\bnfnt{op} &\bnfeq \bnft{+} \bnfalt \bnft{-} \bnfalt \bnft{*} \bnfalt \bnft{/} \\
\bnfnt{num} &\bnfeq \bnft{1} \bnfalt \bnft{2} \bnfalt \bnft{3}
\end{align*}
\caption{BNF specification for arithmetic expressions using many parentheses}
\label{fig:arith-paren}
\end{figure}

Another solution to the ambiguity problem is to surround \textit{every} application of an operator with parentheses.
\Cref{fig:arith-paren} has such a grammar for arithmetic expressions.
%
\begin{exercise}
Derive the sentence \codel{((1 * (2 * (3 - 2))) + 1)} using the grammar in \Cref{fig:arith-paren}.
\end{exercise}
%
The issue with this approach is that it means our expressions has a lot of parentheses.
This can be frustrating, but in reality it's not so bad.
Note that, if we parenthesize everything, the fixity of operators becomes irrelevant; the parentheses play entirely the role of disambiguating multiple applications of operators.
Furthermore, if we design our grammar so that all operators are prefix, then we get \textbf{S-expressions}.
This might feel like piling one hard-to-read syntactic paradigm onto another, but there are quite a few programming languages whose surface-level syntax is made up of S-expressions.\footnote{
  I understand this might not be a compelling argument to the parentheses-intolerant among us (of which there are far too many) but it at least indicates that the multitude of parentheses is an aesthetic \textit{inclination}, and not just a quirk of the bizarre and aged.
}
And the benefits are vast: S-expressions are dead-simple to parse, which makes for quick programming language development.
They also allow for some fancier features we won't go into, but make for convenient macro systems in languages like Lisp and Scheme.\footnote{
  If you're interested, look up the term \textit{homoiconicity}.
}

\begin{remark}
  \label{rmk:parens}
  One common conceptual roadblock is in recognizing that \textit{parentheses are part of syntax}.
  We're trained to think of parentheses as meta-syntax, included only to make things more clear, but never explicitely considered (e.g., when you learned calculus, you likely didn't have a lesson on \enquote{the use of parentheses in analytic expressions}).
  But we're on \enquote{the other side} so to speak, meaning we're the ones who have to specify exactly (i.e., formally) how parentheses work.
  In other words, we don't get parentheses \enquote{for free,} we have to implement them as part of the syntax of the language.
\end{remark}
%
All told, our basic question becomes: \textit{how do we avoid grammatical ambiguity while being able to mix operator fixity and not use so many parentheses?}
And this question has a simple answer in theory: \textit{make explicit assumptions about how operator arguments should be grouped.}
This will mean contending with two things: \textbf{associativity} and \textbf{precedence}.

\subsection{Associativity}

Associativity refers to how arguments are grouped when we're given a sequence of applications of an infix operator in the absence of parentheses.
For example, the expression:

\begin{ocaml}
1 + 2 + 3 + 4
\end{ocaml}
%
can be understood as any one of the following:

\begin{ocaml}
  ((1 + 2) + 3) + 4
  (1 + (2 + 3)) + 4
  (1 + 2) + (3 + 4)
  1 + ((2 + 3) + 4)
  1 + (2 + (3 + 4))
\end{ocaml}
%
In the case of addition, the point is somewhat moot.
The order in which we group arguments doesn't affect the \textit{value} of a sequence of additions.
That is, addition is \textbf{associative}.
%
\begin{definition}
  An operation $\circ : X \to X \to X$ is \textbf{associative} if
  %
  \begin{align*}
    (a \circ b) \circ c = a \circ (b \circ c)
  \end{align*}
  %
  for all $a$, $b$, and $c$ in $X$.
\end{definition}
%
But not all operators are associative, e.g., subtraction.
We need to \textit{decide} how to implicitly parenthesize an expression like \codel{1 - 2 - 3 - 4}.
For binary operators, we typically choose one of the following two ways of grouping terms.
%
\begin{definition}
  An operator $\circ : X \to X \to X$ is said to be \textbf{left-associative} if sequences of applications of the operator are understood as grouping arguments from left to right, i.e.,
  %
  \begin{align*}
    a \circ b \circ c \circ d = (((a \circ b) \circ c) \circ d)
  \end{align*}
  %
  for any $a$, $b$, $c$, and $d$, in $X$.
  It's said to be \textbf{right-associative} if arguments are grouped from right to left, i.e.,
  %
  \begin{align*}
    a \circ b \circ c \circ d = (a \circ (b \circ (c \circ d)))
  \end{align*}
\end{definition}
%
Bringing this back to grammatical ambiguity, giving an implicit parenthesization of a sequence of operators means specifying a "shape" for the corresponding parse tree.
Taking subtraction to be left-associative means that one of the two parse trees for the
sentence \codel{1 - 2 - 3} in the previous section is correct (the first one) and the other is not (the second one).

\begin{exercise}
  \label{exr:fun-type}
  The following is a grammar for function types over the base types \codel{int} and \codel{bool}.
  \begin{align*}
    \bnfnt{fun-type} &\bnfeq
    \bnfnt{base-type} \bnfalt
    \bnfnt{fun-type} \ \bnft{->} \ \bnfnt{fun-type} \bnfalt
    \bnft{(} \ \bnfnt{fun-type} \ \bnft{)} \\
    \bnfnt{base-type} &\bnfeq \bnft{int} \bnfalt \bnft{bool}
  \end{align*}
  %
  Give two leftmost derivation of \codel{int -> bool -> int}.
\end{exercise}
%
All of this rests on the following mathematical fact about the formal grammars.

% https://www.cse.chalmers.se/edu/year/2012/course/TMV026/lec13.pdf
\begin{definition}
  A symbol $X$ is \textbf{useful} in a BNF grammar $\mathcal G$ if there is a derivation of a sentence $S$ in $\mathcal G$ which uses the symbol $X$.
\end{definition}

\begin{proposition}
  \label{prop:bin-ambig}
  Let $\mathcal G$ be a BNF grammar and let $\bnfnt{a}$ and $\bnfnt{b}$ be arbitrary nonterminal symbols which are useful in $\mathcal G$.
  Further suppose that
  \begin{align*}
    \bnfnt{a} \bnfeq \bnfnt{a} \ \bnfnt{b} \ \bnfnt{a}
  \end{align*}
  is a production rule of $\mathcal G$.
  Then $\mathcal G$ is ambiguous.
\end{proposition}

\begin{exercise}
  \textbf{(Challenge)} Prove \Cref{prop:bin-ambig}.
\end{exercise}
%
Generally speaking we should be wary of production rules of the form given in \Cref{prop:bin-ambig}, because it's possible to construct subtrees along the lines of the ones given for subtraction above:

\begin{center}
  \begin{forest}
    [$\bnfnt{a}$
      [$\bnfnt{a}$
        [$\bnfnt{a}$]
        [$\bnfnt{b}$]
        [$\bnfnt{a}$]
      ]
      [$\bnfnt{b}$]
      [$\bnfnt{a}$]
    ]
  \end{forest}
  \qquad
  \begin{forest}
    [$\bnfnt{a}$
      [$\bnfnt{a}$]
      [$\bnfnt{b}$]
      [$\bnfnt{a}$
        [$\bnfnt{a}$]
        [$\bnfnt{b}$]
        [$\bnfnt{a}$]
      ]
    ]
  \end{forest}
\end{center}
%
\Cref{prop:bin-ambig} implies that a grammar for arithmetic will be ambiguous as long as it contains the rule
\begin{align*}
  \bnfnt{expr} \bnfeq \bnfnt{expr} \ \bnfnt{op} \ \bnfnt{expr}
\end{align*}
%
as the grammar in \Cref{fig:arith-simple} does.
The question remains: \textit{what do we do about this?}
According to the discussion above, this is in part an issue of associativity.
The problem arises from the fact that we can expand \textit{both} the left and the right argument of an operator to be another application of a binary operator.

The first step is to specify an associativity assumption for binary operators.
It's typical to take addition, subtraction, multiplication and division to be \textit{left associative}, e.g., this is the \enquote{correct} parse tree for \codel{1 - 2 - 3}:

\begin{center}
  \begin{forest}
    [$\bnfnt{expr}$
      [$\bnfnt{expr}$
        [$\bnfnt{num}$
          [$\bnft{1}$]
        ]
        [$\bnfnt{op}$
          [$\bnft{-}$]
        ]
        [$\bnfnt{num}$
          [$\bnft{2}$]
        ]
      ]
      [$\bnfnt{op}$
        [$\bnft{-}$]
      ]
      [$\bnfnt{num}$
        [$\bnft{3}$]
      ]
    ]
  \end{forest}
\end{center}
%
Another way of expressing left-associativity is that the \textit{right} argument of an operator can't be another application of the operator.
In order to enforce left-associativity of operators, we need to \enquote{factor out} the application of a binary operator.
Said another way, we need to create a \textit{new} nonterminal symbol $\bnfnt{expr-no-op}$ which is the same as $\bnfnt{expr}$ but which can't be expanded to an application of a binary operator.
This is done for the grammar in \Cref{fig:arith-left}.

\begin{figure}
  \begin{align*}
    \bnfnt{expr} &\bnfeq
    \bnfnt{expr} \ \bnfnt{op} \ \bnfnt{expr-no-op} \bnfalt
    \bnfnt{expr-no-op} \\
    \bnfnt{expr-no-op} &\bnfeq
    \bnfnt{num} \bnfalt
    \bnft{(} \ \bnfnt{expr} \ \bnft{)} \\
    \bnfnt{op} &\bnfeq \bnft{+} \bnfalt \bnft{-} \bnfalt \bnft{*} \bnfalt \bnft{/} \\
    \bnfnt{num} &\bnfeq \bnft{1} \bnfalt \bnft{2} \bnfalt \bnft{3}
  \end{align*}
  \caption{BNF specification for arithmetic with left-associative operators}
  \label{fig:arith-left}
\end{figure}

\begin{exercise}
  \label{exr:arith-left}
  Draw parse trees for the following sentences using the grammar in \Cref{fig:arith-left}.
  \begin{align*}
    &\codel{1 - 2 - 3} \\
    &\codel{1 * (2 - 3)} \\
    &\codel{1 + 2 * 3}
  \end{align*}
\end{exercise}

\begin{proposition}
  \label{prop:arith-left-unambig}
  The grammar in \Cref{fig:arith-left} is unambiguous.
\end{proposition}

\begin{exercise}
  \textbf{(Challenge)}
  Prove \Cref{prop:arith-left-unambig}.
\end{exercise}

\begin{exercise}
  Rewrite the grammar in Exercise \ref{exr:fun-type} so that it is unambiguous and the function type arrow is taken to be \textit{right} associative.
\end{exercise}

\subsection{Precedence}

We're getting closer to a grammar that captures our intuitions about arithmetic expressions.
We've succeeded in designing a grammar that is not ambiguous, but as is demonstrated in Exercise \ref{exr:arith-left}, we've failed to enforce an aspect of arithmetic expressions that we learn in grade school: \codel{1 + 2 * 3} should be equivalent to \codel{1 + (2 * 3)} and not \codel{(1 + 2) * 3}.
This is an issue of \textbf{precedence} or \textbf{order of operations}, something that you're likely already familiar with.
And like associativity, incorrectly dealing with precedence can affect the output value of a program.

If you went through the American public school system then you probably learned the abbreviation PEMDAS (Parentheses, Exponentiation, Multiplication, Division, Addition, Subtraction) along with an accompanying mnemonic, something like \enquote{please excuse my dear aunt sally.}
Focusing on just the last four letters, this rule tells us that we should group multiplications and divisions first, and then group additions and subtractions.
That is to say, multiplication and division have greater \textbf{precedence} than addition and subtraction.

\begin{definition}
  The \textbf{precedence} of an operator, relative to another operator, determines which operator binds more tightly, in the presents of ambiguity.
\end{definition}
%
Like associativity, the relative precedence of a collection of operators determines the shape of the parse tree.
To say that multiplication has higher precedence is to say that when we build the parse tree for \codel{1 * 2 + 3}, the operator \enquote{\codel{+}} should be the top-level operation, i.e., at the root of the tree.

\begin{remark}
  \label{rmk:diff-assoc}
  One thing that was probably glossed over if/when you learned PEMDAS:
  \textit{what do you do with something like \codel{1 + 2 - 3 + 4}?}
  Do you group additions and then subtractions?
  Or vice versa?
  The issue here is that addition and subtraction have the \textit{same} precedence (something which is not made clear by PEMDAS).
  In this case, we'll use the associativity of the operators to determine how to parenthesize, i.e., given a sequences of operators of the same precedence, we use their associativity to group them.

  Since addition and subtraction are both left-associative, the above expression should be equivalent to \codel{((1 + 2) - 3) + 4}.
  Things get complicated if we have two operators with the same precedence, but \textit{different} associativity.
  We'll ignore this possibility, but this matters in languages like Haskell, where users can define their own operators with specified precedence and associativity.
\end{remark}
%
How do we deal with precedence?
In the same way that as we dealt with associativity: by \enquote{factoring out} the operators of higher precedence.
Another way of expressing that \enquote{\codel{*}} has higher precedence than \enquote{\codel{+}} is to say that the arguments of \enquote{\codel{*}} can't be applications of \enquote{\codel{+}} (of course we can still put additions in parentheses).
Said another way, we need to create a new nonterminal symbol $\bnfnt{expr-no-pm}$ which is the same as $\bnfnt{expr}$ but which can't be expanded to an addition or subtraction.
This is done for the grammar in \Cref{fig:arith-final}.
Note that, in this grammar, we also maintain the asymmetry introduced to deal with associativity.

\Cref{fig:arith-final-tree} has a parse tree for the sentence \codel{1 + 2 * 3} in this grammar.
Naturally the tree is \enquote{noisier} in that there's more intermediate nonterminal symbols, but hopefully the point is clear: we have to expect with addition first because otherwise, we'd get stuck.
That is, if we expand the multiplication first, the grammar disallows us expanding the left argument to \codel{1 + 2}.

\begin{figure}
  \begin{align*}
    \bnfnt{expr} &\bnfeq
    \bnfnt{expr} \ \bnfnt{pm} \ \bnfnt{expr-no-pm} \bnfalt
    \bnfnt{expr-no-pm} \\
    \bnfnt{expr-no-pm} &\bnfeq
    \bnfnt{expr-no-pm} \ \bnfnt{md} \ \bnfnt{expr-no-op} \bnfalt
    \bnfnt{expr-no-op} \\
    \bnfnt{expr-no-op} &\bnfeq
    \bnfnt{num} \bnfalt
    \bnft{(} \ \bnfnt{expr} \ \bnft{)} \\
    \bnfnt{pm} &\bnfeq \bnft{+} \bnfalt \bnft{-} \\
    \bnfnt{md} &\bnfeq \bnft{*} \bnfalt \bnft{/} \\
    \bnfnt{num} &\bnfeq \bnft{1} \bnfalt \bnft{2} \bnfalt \bnft{3}
  \end{align*}
  \caption{BNF specification for arithmetic with left-associative operators}
  \label{fig:arith-final}
\end{figure}

\begin{figure}
  \begin{forest}
    [$\bnfnt{expr}$
      [$\bnfnt{expr}$
        [$\bnfnt{expr-no-pm}$
          [$\bnfnt{expr-no-op}$
            [$\bnfnt{num}$
              [$\bnft{1}$]
            ]
          ]
        ]
      ]
      [$\bnfnt{pm}$
        [$\bnft{+}$]
      ]
      [$\bnfnt{expr-no-pm}$
        [$\bnfnt{expr-no-pm}$
          [$\bnfnt{expr-no-op}$
            [$\bnfnt{num}$
              [$\bnft{2}$]
            ]
          ]
        ]
        [$\bnfnt{md}$
          [$\bnft{*}$]
        ]
        [$\bnfnt{expr-no-op}$
          [$\bnfnt{num}$
            [$\bnft{3}$]
          ]
        ]
      ]
    ]
  \end{forest}
  \centering
  \caption{Parse tree of \codel{1 + 2 * 3} using the grammar in \Cref{fig:arith-final}}
  \label{fig:arith-final-tree}
\end{figure}

\begin{exercise}
  Give a derivation of \codel{(1 + 2) * 3} using the grammar in \Cref{fig:arith-final}
\end{exercise}

\subsection{Operator Tables}
In the next chapter, we'll see how to implement parsers for a given grammar.
One way is to use a \textbf{parser generator}, which is a program that generates the code for a parser based on a specification.
More on that later, the point: most parser generators don't require us to use the techniques above to deal with associativity and precedence.
Rather, they take additional about associativity and precendence directly, often in the form of an \textbf{operator table}.
For arithmetic, we might use the following table, which presents operators in order of increasing precedence.

\begin{center}
  \begin{tabular}{|c|c|}
    \hline
    Operator & Associativity \\
    \hline
    \codel{+}, \codel{-} & left \\
    \codel{*}, \codel{/} & right \\
    \hline
  \end{tabular}
\end{center}
%
Note that, in light of \Cref{rmk:diff-assoc}, it's reasonable to assume that all operators of the same precedence (i.e., appearing in the same row) have the same associativity.
The \href{https://ocaml.org/manual/5.3/api/Ocaml_operators.html}{OCaml Manual} has an analogous table for its operators.

\begin{remark}
  We can think of this as an \textit{a posteriori} approach to disambiguation; rather than designing a grammar in which every sentence has a unique parse tree, we design an ambiguous grammar and then specify which of the parse trees we consider acceptable.
  As long as the only source of ambiguity is due to associativity and precedence of binary operators, operator tables allow use to associate with every sentence a unqiue parse tree.
\end{remark}

A final word of warning for this section: as much as we've tried to present grammars as a rigorously as possible, they tend to be a bit handwavy in practice.
For example, the following is a standard presentation of arithmetic expressions in a more mathematical setting:
\begin{align*}
  e \bnfeq e + e \bnfalt e - e \bnfalt e * e \bnfalt e / e \bnfalt \mathbb Z
\end{align*}
where $\mathbb Z$ refers to any integer.
Operator associativity and precedence are understood from context and the production rule:
\begin{align*}
  e \bnfeq ( e )
\end{align*}
is implicitly assumed, contrary to \Cref{rmk:parens}!
In other settings, (e.g., the \href{https://docs.python.org/3/reference/grammar.html}{full grammar specification} of Python) the notation used is a mixture of multiple syntaxes for grammar specifications.
This is just to say, there is a difference between grammars as they are studied in formal language theory, and as they are used to specify programming languages or used to specify syntaxes in mathematical settings like logic and type theory.
In the later cases, you'll likely have to read between the lines, so to speak.

\section{Extended BNF}

\section{Regular Expressions}

\section{Grammars as Formal Systems}
