When we write a program in our favorite programming language, we fill a file with a collection of symbols that we typed up using a computer keyboard.
This is one of the beauties of programming: looking past the bells and whistles provided by editors and programming tools, a program is just a stream of characters.
At some point in our programming workflow, we need to verify that what we've typed up works as expected, so we \textit{run} our program.
In some editors, there is literally a \enquote{play} button, but in many other cases this means opening a terminal and typing out a few commands.
In either case, we're running a \textit{different} program in order to run the program we've written.
Our goal is to understand what's going on here:
\textit{
  What is this program doing?
  How does it do it?
  What sorts of data structures does this program use?
}
Our intuition should tell us this program is probably doing something fairly complicated.\footnote{Just think of how hard it is to follow directions, and how often we wish we read \textit{all} the instructions before starting some task.}
So our basic question is: \textit{How do we get from a stream of symbols representing a program to its output?}

\section{The Interpretation Pipeline}

The program to which we referred in the previous paragraph\textemdash the one that runs the programs we write\textemdash is called an \textit{interpreter}.
An interpreter takes as input a stream of characters, along with additional inputs (perhaps given as command line arguments or environment variables) and produces the output of our program.
So another way of phrasing our basic question is simply: \textit{what does an interpreter do?}
As is typically the case, it's easier to answer a question like this by breaking it up into several subquestions.
These subquestions correspond to passes in the \textit{interpretation pipeline}, which is visualized in \Cref{fig:interp-pipe}.  As we will understand it, an interpreter does four things:
\begin{enumerate}
\item
  It attempts to convert a sequence of characters (our program) into a sequence of \textit{tokens}.
  We can think of tokens as the \textit{units} or \textit{atoms} of our program.
  This part of the interpreter is called \textbf{lexical analysis}, and its primary purpose is to simplify the process of analyzing the input program.
  It's easier, for example, to analyze a program if we know beforehand whether the character \codel{1} which appears in our program is part of a number, a variable name, or a comment.
  Lexical analysis handles all these low-level syntactic concerns up front.
  In analogy with natural language, this is like the part of language processing in which your brain combines sequences of sounds or letters into whole words.
\item
  Once we have a sequence of tokens, our interpreter will attempt to convert them into an \textit{abstract syntax tree (AST)}, which is a representation of our program as \textit{hierarchical data}.
  This part of the interpreter is called \textbf{syntactic analysis} or \textbf{parsing}.
  Hierarchical data is easier to analyze and evaluate, e.g., we'd like to know before evaluating our program if a variable \codel{x} that might appear in our program is part of the expression we're evaluating or part of a new variable declaration.
  In analogy with natural language, this is like when your brain combines words into sub-phrases like prepositional phrases and verb phrases.
\item
  Not all programs we can write make sense.
  It doesn't make sense, for example, to add a number to a string.\footnote{Even though some languages allow you to do this.}
  The next part of an interpreter should analyze the hierarchical data from the previous part to verify that everything in it \enquote{looks reasonable.}
  There are a couple ways to do this; this is generally a part of what is called \textbf{static (semantic) analysis}, but one of the most common form\textemdash and the way on which we'll focus because it's OCaml's approach\textemdash is \textbf{type checking}.
  Types help us describe what kind of things we're working with \textit{before we start running our program}.
  They help us determine \textit{before we start running our program} if we're working with data correctly.
  In analogy with natural language, this is like the uneasy feeling you get when you hear or see the sentence \enquote{the happiness is cold.}
  The word \enquote{happiness} is not the right category of word for its location in the sentence, despite the fact that this sentence is grammatically correct.
  Not all languages have type checkers, but we maintain that \enquote{good} languages have type checkers.
\item
  Everything above is in service of \textit{running} our program.
  This part of the interpreter is called \textbf{dynamic (semantic) analysis} or \textbf{evaluation}.
  This is the most intuitive part of an interpreter for programmers; learning to program is ultimately internalizing the evaluation rules of a programming language, so that we know what to write to accomplish a certain task.
\end{enumerate}
This can be seen as our roadmap; we will consider each of these of parts in turn.
For each part, there will be two perspectives: the theoretical perspective and the practical perspective.
For example, the formal counterpart of syntactic analysis is \textbf{formal grammar} and the practical counterpart is \textbf{parsing}.
We will build simple interpreters along the way, and by the end of the text we'll (hopefully) have an appreciation for the kinds of considerations that go into designing \enquote{good} programming languages.

\begin{remark}
It should be noted at the point that we will \textit{not} consider \textbf{compilation}.
In rough terms, compilation is the process of translating a program into another program in some target language, often a language that deals with low-level concerns like memory management and hardware architecture (e.g., assembly language or LLVM).
We will always consider high-level semantics for our programming languages.
And when we implement interpreters, we will work in high-level abstractions using OCaml.
\end{remark}


\begin{figure}
  \hspace{-1.5cm} % ugh
  \begin{tikzpicture}[
      interpbox/.style={
        rectangle,
        rounded corners,
        draw,
        thick,
        minimum width=6cm,
        minimum height=1cm,
      }
    ]
    \draw[thick, rounded corners] (2,1) rectangle (9, -12.25);
    \node (chars) {\textit{stream of characters}};
    \node[interpbox] (lex) [right=of chars] {Lexical Analysis};
    \node (tokens) [below=of lex] {\textit{stream of tokens}};
    \node[interpbox] (par) [below=of tokens] {Syntactic Analysis (Parsing)};
    \node (ast) [below=of par] {\textit{abstract syntax tree (AST)}};
    \node[interpbox] (sem) [below=of ast] {Semantic Analysis (Type Checking)};
    \node (ir) [below=of sem] {\textit{well-typed AST/intermediate repr. (IR)}};
    \node[interpbox] (eval) [below=of ir] {Evaluation};
    \node (input) [left=of eval] {\textit{input}};
    \node (output) [right=of eval] {\textit{output}};
    \draw[->, thick] (chars.east) -- (lex.west);
    \draw[thick] (lex.south) -- (tokens.north);
    \draw[->, thick] (tokens.south) -- (par.north);
    \draw[thick] (par.south) -- (ast.north);
    \draw[->, thick] (ast.south) -- (sem.north);
    \draw[thick] (sem.south) -- (ir.north);
    \draw[->, thick] (ir.south) -- (eval.north);
    \draw[->, thick] (input.east) -- (eval.west);
    \draw[->, thick] (eval.east) -- (output.west);
    \node at (7.7, -12) {\texttt{INTERPRETER}};
  \end{tikzpicture}
  \centering
  \caption{
    Visualization of the interpretation pipeline
  }
  \label{fig:interp-pipe}
\end{figure}
