This text is by its nature a bit strange.
The scope is quite limited: it's the source text for the second half of \textbf{CAS CS 320: Principles of Programming Languages} at Boston University, and the intended audience is ultimately students in this course.
My hope is that it might find a wider audience, but describing a more inclusive notion of \enquote{intended audience} requires some context: CS320 is a course in the computer science cirriculum at Boston University which is semi-required.\footnote{I won't go into the details of the \enquote{semi} part here.}
As much as I appreciate this, and feel that everyone should know something about how programming languages work, there is a non-negligible slice of the student population which passes through our course who don't agree with this sentiment.
And so we're put in the difficult but perennial position: \textit{how to we get students interested in a topic they're studying because they're told, in a sense, they have to?}
Getting to the point: the intended audience of this text is undergraduate students who don't necessarily want to do research in programming languages, but \enquote{want} to know something about how programming languages work.
There's a lot we don't cover, and although we try to maintain the rigor of some of the more serious texts in the field \cite{} (all of which are absolutely brilliant), we're not gonna to lose sleep over the less-than-complete coverage we inevitably end up with.

I glossed over this point above, but another operative part of the description of our context of is \enquote{the second half of}.
We typically spend the first half of the course teaching OCaml from TODO.
In fact, should in some sense be viewed as an extension of this text, with section on interpreters fleshed out, and as such we're greatly in debt to this lovely work.

The last point I'll make is simply that this text is strange because I myself wrote it.
I don't really want to claim any role in the bizarre nature of this text, but the truth is I don't know how to write a wholly \enquote{normal} text.
I come to programming language from the perspective of logic and type theory, which colors the topics covered therein, but I also have been fed on a steady diet of logicians and mathematicians who are clever writers in so far as they really only know how to write math (e.g., Raymond Smullyan, Girard etc.) or writers who are actually clever writers but who don't know how to math (e.g., David Foster Wallace), the tex formatting obsessed (TODO).
I'd like to believe I lie somewhere in the middle, hopefully not to \enquote{writerly} or \enquote{mathy} but in a way that in some sense I cannot avoid it.

This text is based on material
They are based on material created by several members of the \textit{Principles of Programming and Verification (POPV)} group at Boston University, most notably Marco Gaboardi, Hongwei Xi, Robin, Ankush Das, and Assaf Kfoury.
